From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-21 paper/paper.Rmd
WIP

What is noise? Where does it come from?  Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise [e.g. @Coulson2004]. This literature
has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intriniscally 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models.

@@

; @Ovaskainen2010 the canonical on the highly influential @Leigh1981, @Lande1993, @Foley1994

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise, and unsurprisingly few topics have
recieved greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

>>>>

What is noise? Where does it come from? 
Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level in deterministic settings. The notion of a deterministic
"skeleton" -- responsible for the phenomena of interest -- that is obscured
from view by the nusiance of noise has slowly given way to recognition
that stochastic and deterministic elements not only interact in 
interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise.

Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
This literature has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intriniscally 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

In the extensive literature developing the consequences of both demographic
and environmental noise in models of increasing complexity, it can 
be easy to miss the equally important developments in making the origins
of these sources of noise more precise -- developments which have important
consquences for when and how the resulting models can be applied.

the canonical model on the highly influential @Leigh1981, @Lande1993, @Foley1994

$$ \frac{\textrm{d} n}{\textrm{d} t} = \underbrace{f(n)}_{\textrm{det skeleton}} + \underbrace{\sigma_d \xi_d(t) \sqrt{n}}_{\textrm{demographic noise}} + \underbrace{\sigma_e \xi_e(t) n}_{\textrm{environmental noise}}$$

Stochastic differential equations of this form are so-called "Langevin equation," popular in the physics literature.  This notation is rather sloppy, as the standard calculus does not apply to stochastic equations.  The Langevin formulation, most popular in physics, is usually associated with a Stratonovich integral.  Mathematicians prefer to write stochastic differential equations as $\textrm{d} n = f(x) \textrm{d} t + g(x) \textrm{d} B_t$, where the differentials hint at the underlying definition that this is the sum of a classic Reimann integral (over $\textrm{d}t$) and a stochastic It\^o integral.

Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models.

@@

; @Ovaskainen2010

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].

*Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise, and unsurprisingly few topics have
recieved greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.
